An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-05-23 full_paper_v2.tex
initial

@@

This paper adopts an interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, an
d reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

@@

While the psychosocial factors proposed in Williams and Andersen's
(Williams and Andersen, 1998) model have received the most research
attention, the mechanisms through which these factors are proposed to
exert their effect have remained under-investigated in the literature.
To elaborate, the model suggests that injuries are likely to occur
through either increased physiological arousal resulting in increased
muscle tension and reduced flexibility or attentional deficits caused by
increased distractibility and peripheral narrowing. However, to date,
the research has largely focused on attentional deficits (Andersen and
Williams, 1999; Rogers and Landers, 2005; Wilkerson, 2012; Swanik et
al., 2007). For example, Andersen and Williams (Andersen and Williams,
1999) measured peripheral and central vision during high and low stress
conditions and found athletes with high life event stress coupled with
low social support had greater peripheral narrowing under stressful
conditions compared to athletes with the opposing profile; these
athletes went on to sustain an increased number of injuries during the
following season. Indeed, Rodgers and Landers (Rogers and Landers, 2005)
supported Andersen and Williams's (Andersen and Williams, 1999) earlier
findings reporting that peripheral narrowing under stress mediated 8.1\%
of the relationship between negative life events and injury. However,
the remaining variance between negative life events and athletic injury
through the other proposed mechanisms, such as increased muscle tension
and reduced motor control, remains to be explored (cf.~Williams and
Andersen, 1998).

One challenge faced by researchers addressing the sports injury problem
with a psychological lens is the multifactorial nature of injury, and
the possible contribution of other non-psychological factors to the
stress response (Meeuwisse et al., 2007; Wiese-Bjornstal, 2009). For
example, a large body of research indicates that training-related stress
is also likely to be related to the stress response and injury
occurrence (Lee et al., 2017; Djaoui et al., 2017), and may account for
the unexplained variance from the psychological predictors of injury.
Appaneal and Perna (Appaneal and Perna, 2014) proposed the
biopsychosocial model of stress athletic injury and health (BMSAIH) to
serve as an extension to Williams and Andersen's (Williams and Andersen,
1998) model and to address some of these issues. To elaborate, the
BMSAIH aimed to clarify the mediating pathways between the stress
response and injury, consider other health outcomes and behavioural
factors that impact sports participation, and integrate the impact of
training on athletes' health (Appaneal and Perna, 2014). The central
tenet of the BMSAIH is that psychosocial distress (e.g., negative life
events) may act synergistically with training-related stress as a result
of high-intensity and high-volume sports training, and ``widen the
window of susceptibility'' (Appaneal and Perna, 2014, 74) to a range of
undesirable health outcomes including illness and injury. Consequently,
the BMSAIH provides a framework for future research to build on Williams
and Andersen's (Williams and Andersen, 1998) model, by including other
physiological markers of training-related stress, which together may
provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the
relationship between hormonal responses to training and injury
occurrence (Perna and McDowell, 1995; Perna et al., 1997, 2003), other
research has identified additional markers of training-related stress
that are associated with an increased risk of injury; for example, heart
rate variability (Bellenger et al., 2016; Williams et al., 2017),
postural stability (Romero-Franco et al., 2014) and muscle stiffness
(Pruyn et al., 2015). However, these markers are often studied in
isolation without an assessment of the psychosocial factors that are
known to contribute to injury, thereby limiting our understanding of how
psychosocially and physiologically derived stress may contribute
synergistically to injury occurrence. Recently, Bittencourt et
al.~(Bittencourt et al., 2016) suggested that to better understand the
multifactorial nature of sports injuries, research needs to move away
from studying risk factors in isolation and instead adopt a complex
systems approach to injury. Such an approach posits that injury may
arise from a complex ``web of determinants'' (Bittencourt et al., 2016,
3), where different factors interact in unpredictable and unplanned
ways, but result in a global outcome pattern of either adaptation or
injury.

A challenge when adopting a complex systems approach is using an
appropriate analysis technique that is able to capture the uncertainty
and complexity of the relationships between different factors. One
technique that provides a solution to this challenge is Bayesian network
(BN) modelling. BN's allow the construction of graphical probabilistic
models using the underlying structure that connects different variables
(nodes in the network) of interest (Scutari and Denis, 2014). Once
constructed, the learned structure can be used for inference by
obtaining the posterior probabilities of a particular node for a given
query (e.g., if the value of Node A is x and the value of Node B is y,
what is the probability Node C being value z?). Furthermore, BN's do not
distinguish between dependent and independent variables as they are a
form of unsupervised learning, which is a strength over regression or
structure equation models when the underlying relationship in the
network may not be known (Olmedilla et al., 2018).

To summarise, despite offering a possible framework to build on the
research stemming from Williams and Andersen's (Williams and Andersen,
1998) model, there remains and opportunity to explore other
physiological stress-related markers proposed by the BMSAIH, in addition
to the already well-established psychological characteristics known to
be related to injury (Appaneal and Perna, 2014). Furthermore, research
has typically not captured changes in both psychosocial factors and
stress-related physiological markers that may occur between initial
measurement and injury occurrence. Given the exploratory the current
study had the following objectives:

\begin{itemize}
\tightlist
\item
  Identify suitable markers of stress that can be easily captured in a
  large cohort of athletes in a field based setting.
\item
  Examine the relationships between the markers of stress and injury
  using a prospective, repeated measures design.
\item
  Evaluate the relationships between the markers of stress and injury
  using Bayesian network modelling that captures the complex nature of
  injury occurrence.
\end{itemize}

@@

@@

\leavevmode\hypertarget{ref-Olmedilla2018}{}%
Olmedilla, A., Rubio, V. J., Fuster-Parra, P., Pujals, C., and
García-Mas, A. (2018). A Bayesian approach to sport injuries likelihood:
Does player's self-efficacy and environmental factors plays the main
role? \emph{Frontiers in Psychology} 9, 1--10.
doi:\href{https://doi.org/10.3389/fpsyg.2018.01174}{10.3389/fpsyg.2018.01174}.

@@

@@

\leavevmode\hypertarget{ref-Scutari2014}{}%
Scutari, M., and Denis, J.-B. (2014). \emph{Bayesian networks: with
examples in R}. 1st ed. Chapman \& Hall/CRC Available at:
\url{https://www.crcpress.com/Bayesian-Networks-With-Examples-in-R/Scutari-Denis/p/book/9781482225587}.

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

This paper adopts an interdisciplinary approach to explore the relationship between psychosocial factors, physiological stress-related markers and occurrence of injury in athletes using a repeated measures design across a 2-year data collection period. At three data collection time-points, athletes completed measures of major life events, the reinforcement sensitivity theory personality questionnaire, muscle stiffness, heart rate variability and postural stability, and reported any injuries they had sustained since the last data collection. Two Bayesian networks were used to examine the relationships between variables and model the changes between data collection points in the study. Findings revealed muscle stiffness to have the strongest relationship with injury occurrence, with high levels of stiffness increasing the probability of sustaining an injury. Negative life events did not increase the probability of injury occurrence at any single time-point; however, when examining changes between time points, increases in negative life events did increase the probability of injury. In addition, the combination of increases in negative life events and muscle stiffness resulted in the greatest probability of sustaining an injury. Findings demonstrated the importance of both an interdisciplinary approach and a repeated measures design to furthering our understanding of the relationship between stress-related  markers and injury occurrence.

@@

While the psychosocial factors proposed in Williams and Andersen (1998)
have received

Research that has explored the effects of the psychosocial factors in
Williams and Andersen's (Williams and Andersen, 1998) model, has failed
to examine the mechanisms through which these factors are proposed to
exert their effect. To elaborate, the model suggests that injuries are
likely to occur through either increased physiological arousal resulting
in increased muscle tension and reduced flexibility or attentional
deficits caused by increased distractibility and peripheral narrowing.
However, to date, the research has largely focused on attentional
deficits (Andersen and Williams, 1999; Rogers and Landers, 2005;
Wilkerson, 2012; Swanik et al., 2007). For example, Andersen and
Williams (Andersen and Williams, 1999) measured peripheral and central
vision during high and low stress conditions and found athletes with
high life event stress coupled with low social support had greater
peripheral narrowing under stressful conditions compared to athletes
with the opposing profile; these athletes went on to sustain an
increased number of injuries during the following season. Indeed,
Rodgers and Landers (Rogers and Landers, 2005) supported Andersen and
Williams's (Andersen and Williams, 1999) earlier findings reporting that
peripheral narrowing under stress mediated 8.1\% of the relationship
between negative life events and injury. However, few attempts have been
made to explain the remaining variance between negative life events and
athletic injury through the other proposed mechanisms, such as increased
muscle tension and reduced motor control (cf.~Williams and Andersen,
1998).

One possible reason for this oversight is the multifactorial nature of
injury and the possible contribution of other non-psychological factors
to the stress response (Meeuwisse et al., 2007; Wiese-Bjornstal, 2009).
For example, a large body of research indicates that training-related
stress is also likely to be related to the stress response and injury
occurrence (Lee et al., 2017; Djaoui et al., 2017), and may account for
the unexplained variance from the psychological predictors of injury.
Appaneal and Perna (Appaneal and Perna, 2014) proposed the
biopsychosocial model of stress athletic injury and health (BMSAIH) to
serve as an extension to Williams and Andersen's (Williams and Andersen,
1998) model and to address some of these issues. To elaborate, the
BMSAIH aimed to clarify the mediating pathways between the stress
response and injury, consider other health outcomes and behavioural
factors that impact sports participation, and integrate the impact of
training on athletes' health (Appaneal and Perna, 2014). The central
tenet of the BMSAIH is that psychosocial distress (e.g., negative life
events) may act synergistically with training-related stress as a result
of high-intensity and high-volume sports training, and ``widen the
window of susceptibility'' (Appaneal and Perna, 2014, 74) to a range of
undesirable health outcomes including illness and injury. Consequently,
the BMSAIH provides a framework for future research to build on Williams
and Andersen's (Williams and Andersen, 1998) model, by including other
physiological markers of training-related stress, which together may
provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the
relationship between hormonal responses to training and injury
occurrence (Perna and McDowell, 1995; Perna et al., 1997, 2003), other
research has identified additional markers of training-related stress
that are associated with an increased risk of injury; for example, heart
rate variability (Bellenger et al., 2016; Williams et al., 2017),
postural stability (Romero-Franco et al., 2014) and muscle stiffness
(Pruyn et al., 2015). Unfortunately, these markers are often studied in
isolation without an assessment of the psychosocial factors that are
known to contribute to injury, thereby limiting our understanding of how
psychosocially and physiologically derived stress may contribute
synergistically to injury occurrence. Recently, Bittencourt et
al.~(Bittencourt et al., 2016) suggested that to better understand the
multifactorial nature of sports injuries, research needs to move away
from studying risk factors in isolation and instead adopt a complex
systems approach to injury. Such an approach posits that injury may
arise from a complex ``web of determinants'' (Bittencourt et al., 2016,
3), where different factors interact in unpredictable and unplanned
ways, but result in a global outcome pattern of either adaptation or
injury. Complex systems approaches have been used in health care to
model the large number of risk factors associated with different types
of diseases (Plsek and Greenhalgh, 2001); however, very few studies have
attempted to address sport injury occurrence using such an approach
(Hulme et al., 2018).

To summarise, despite offering a possible framework to build on the
research stemming from Williams and Andersen's (Williams and Andersen,
1998) model, researchers have largely overlooked the potential to
explore other physiological stress-related markers proposed by the
BMSAIH, in addition to the already well-established psychological
characteristics known to be related to injury (Appaneal and Perna,
2014). Furthermore, research has typically not captured changes in both
psychosocial factors and stress-related physiological markers that may
occur between initial measurement and injury occurrence. Viewed through
the lens of a complex systems approach the interaction between
psychosocial sources of stress, stress-related physiological markers and
injury occurrence may provide new insight into the injury process.
Therefore, the purpose of the current study was to examine the
relationship between psychosocial factors, physiological stress-related
markers and occurrence of injury in athletes over time.

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

\hypertarget{subsection-1}{%
\subsection*{Subsection 1}\label{subsection-1}}
\addcontentsline{toc}{subsection}{Subsection 1}

You can use \texttt{R} chunks directly to plot graphs.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\DecValTok{0}\OperatorTok{:}\DecValTok{100}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{999}\NormalTok{)}
\NormalTok{y <-}\StringTok{ }\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{(x }\OperatorTok{+}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\KeywordTok{length}\NormalTok{(x), }\DataTypeTok{sd =} \DecValTok{3}\NormalTok{) }\OperatorTok{+}\StringTok{ }\DecValTok{3}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(x, y)}
\end{Highlighting}
\end{Shaded}

\hypertarget{subsection-2}{%
\subsection*{Subsection 2}\label{subsection-2}}
\addcontentsline{toc}{subsection}{Subsection 2}

Frontiers requires figures to be submitted individually, in the same
order as they are referred to in the manuscript. Figures will then be
automatically embedded at the bottom of the submitted manuscript. Kindly
ensure that each table and figure is mentioned in the text and in
numerical order. Permission must be obtained for use of copyrighted
material from other sources (including the web). Please note that it is
compulsory to follow figure instructions. Figures which are not
according to the guidelines will cause substantial delay during the
production process.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{disclosureconflict-of-interest-statement}{%
\section*{Disclosure/Conflict-of-Interest
Statement}\label{disclosureconflict-of-interest-statement}}
\addcontentsline{toc}{section}{Disclosure/Conflict-of-Interest
Statement}

The authors declare that the research was conducted in the absence of
any commercial or financial relationships that could be construed as a
potential conflict of interest.

\hypertarget{author-contributions}{%
\section*{Author Contributions}\label{author-contributions}}
\addcontentsline{toc}{section}{Author Contributions}

The statement about the authors and contributors can be up to several
sentences long, describing the tasks of individual authors referred to
by their initials and should be included at the end of the manuscript
before the References section.

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgments}

Funding:

\hypertarget{supplemental-data}{%
\section{Supplemental Data}\label{supplemental-data}}

Supplementary Material should be uploaded separately on submission, if
there are Supplementary Figures, please include the caption in the same
file as the figure. LaTeX Supplementary Material templates can be found
in the Frontiers LaTeX folder

\hypertarget{references}{%
\section{References}\label{references}}

A reference list should be automatically created here. However it won't.
Pandoc will place the list of references at the end of the document
instead. There are no convenient solution for now to force Pandoc to do
otherwise. The easiest way to get around this problem is to edit the
LaTeX file created by Pandoc before compiling it again using the
traditional LaTeX commands.

\hypertarget{figures}{%
\section*{Figures}\label{figures}}
\addcontentsline{toc}{section}{Figures}

\begin{figure}

{\centering \includegraphics[width=85mm,height=85mm]{full_paper_v2_files/figure-latex/Figure-1-1}

}

\caption{Figure caption}\label{fig:Figure-1}
\end{figure}

@@

\leavevmode\hypertarget{ref-Hulme2018}{}%
Hulme, A., Thompson, J., Nielsen, R. O., Read, G. J. M. M., and Salmon,
P. M. (2018). Towards a complex systems approach in sports injury
research: Simulating running-related injury development with agent-based
modelling. \emph{British Journal of Sports Medicine} 53, 560--569.
doi:\href{https://doi.org/10.1136/bjsports-2017-098871}{10.1136/bjsports-2017-098871}.

@@

@@

\leavevmode\hypertarget{ref-Plsek2001}{}%
Plsek, P. E., and Greenhalgh, T. (2001). The challenge of complexity in
health care. \emph{BMJ} 323, 625--628.
doi:\href{https://doi.org/10.1136/bmj.323.7313.625}{10.1136/bmj.323.7313.625}.

@@
